---
title: "HW_5.Rmd"
author: "Kira Muzafarov"
date: "2025-04-03"
output: 
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
# Importing Data
Walmart_Sales <- read.csv("/Users/kiramuzafarov/Downloads/Walmart_Sales.csv", header = TRUE)

# Converting Date column to "day-month-year" format.
Walmart_Sales$Date <- as.Date(Walmart_Sales$Date, format = "%d-%m-%Y")

# Extracting year and month.
Walmart_Sales$YearMonth <- format(Walmart_Sales$Date, "%Y-%m")

# Creating a dataframe to fill with monthly sales.
unique_year_months <- unique(Walmart_Sales$YearMonth)
monthly_sales <- data.frame(YearMonth = unique_year_months, Monthly_Sales = rep(NA, length(unique_year_months)))

# Looping through each month and summing up the weekly sales.
for (i in seq_along(unique_year_months)) {current_month <- unique_year_months[i]
  monthly_sales$Monthly_Sales[i] <- sum(Walmart_Sales$Weekly_Sales[Walmart_Sales$YearMonth == current_month], na.rm = TRUE)}

# Converting YearMonth back to Date format (first day of each month)
monthly_sales$YearMonth <- as.Date(paste0(monthly_sales$YearMonth, "-01"))

# Converting to a time series (monthly data, frequency = 12)
start_year <- as.numeric(format(min(monthly_sales$YearMonth), "%Y"))
start_month <- as.numeric(format(min(monthly_sales$YearMonth), "%m"))
ts_monthly_sales <- ts(monthly_sales$Monthly_Sales, start = c(start_year, start_month), frequency = 12)

# Checking structure
str(Walmart_Sales)
summary(Walmart_Sales)

# Loading libraries
library(forecast)
library(ggplot2)
library(fpp)
library(fpp2)
library(forecast)
```

```{r}
# 1. ARIMA Model
ndiffs(ts_monthly_sales)
tsdisplay(ts_monthly_sales)
ts_monthly_salesdiff1 <- diff(ts_monthly_sales, differences=1)
plot(ts_monthly_salesdiff1)
tsdisplay(ts_monthly_salesdiff1)
arima_model <- auto.arima(ts_monthly_sales, stepwise = FALSE, approximation = FALSE)
summary(arima_model)

# The selected ARIMA model is (1,0,0)(0,1,0)[12]. For the non-seasonal part, it uses 1 autoregressive term (AR1), no differencing, and no moving average terms. For the seasonal part, the model applies 1 seasonal difference to make the series stationary and captures yearly seasonality with a period of 12. So the model uses last month’s value and adjusts for seasonal patterns repeating every 12 months, like holiday trends.

# 2. Residual Analysis
checkresiduals(arima_model)
Acf(residuals(arima_model))
Box.test(residuals(arima_model), lag = 20, type = "Ljung-Box")
hist(residuals(arima_model))
tsdiag(arima_model)

# Residuals appear scattered around zero and approximately normally distributed, with no significant autocorrelation. The Ljung-Box test (p = 0.6422) also shows no significant autocorrelation in the residuals, meaning the model fits the data well. The histogram looks fairly normal.

# 3. Forecast Next 5 Months
forecast_result <- forecast(arima_model, h = 5)
autoplot(forecast_result)

#The model appears to pick up on the yearly patterns, which is why the forecast follows a similar seasonal trend.

# 4. Accuracy of the ARIMA Model
accuracy(arima_model)

# Overall the model fits well. The MAPE is 4.60%, which means the model is fairly accurate. This means that on average arima forecasts are within about 4.6% of actual sales. The MASE is 0.51, which is less than 1, indicating that the model performs well. The ACF1 value is -0.21, showing that there’s no strong autocorrelation left in the residual.